---
title: "STATS302 Assignment 1"
author: "Brittany Alexandra"
date: "Wednesday 15 March 2023`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


> 1. 
> -   Perform principal components analysis on the scaled data.
> -   What proportion of the variability is accounted for by the first
>     and second components?

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

#Read data
data<- read.csv("metab.selected.csv")

#Clean data
data<-data[,-c(1:3)]
data<-log(data)
missing<-apply(is.na(data),2 , sum)
data<-data[,missing==0]
sd<-apply(data, 2, sd)
data<-data[,sd>0.001]


#Data matrix 
PCA<-prcomp(data, center=T, scale = T)
plot(PCA)

#Proportion of varaibility
datavaria<-round(PCA$sdev[1:2]^2/sum(PCA$sdev^2),3)
datavaria
```

28.6% of the variability in blood metabolites following Whey Protein
consumption is accounted for by the first principal component, and 9.5%
for the second PC.

In PCA, each principal component represents a linear combination of the
original variables. The first PC captures the most significant patterns
of variation in the data, while the subsequent PCs capture the remaining
variability in decreasing order of importance.

As the first principal component explains 28.6% of the total
variability, we can say it is responsible for a substantial portion of
the patterns observed in the blood metabolites following Whey Protein
consumption. This component likely represents the primary factor
influencing the changes in metabolite levels.

Similarly, the second principal component accounts for 9.5% of the
variability. While it explains less variability compared to the first
PC, it still captures a notable proportion of the patterns in the data.
The second component likely represents additional factors that
contribute to the variability in blood metabolites, possibly
representing specific metabolic responses or interactions.

### 


> 2. Using the rotation matrix and the standard deviation associated with principal component 1 compute the correlation of the first principal component score with the original variables

```{r}
prcomp(data, scale=TRUE)->PCA
par(mfrow=c(2,2))
plot(PCA$rotation[,1], sd, xlab='new PC1', ylab='old PC1') #coefficients in first pc 

coef1<-round(PCA$rotation[,1] * PCA$sdev[1], 2)
sort(coef1, decreasing = TRUE)
```

Positive correlation: A positive correlation indicates that as the PC1
score increases, the corresponding metabolite tends to increase as well.
These metabolites are positively associated with PC1 and may contribute
to its variation.

Negative correlation: A negative correlation suggests that as the PC1
score increases, the corresponding metabolite tends to decrease. These
metabolites are negatively associated with PC1 and may have an inverse
relationship with its variation.

Therefore :

> 3. An advantage of whey protein is that it contains "essential amino acids" (which must come from the diet). The measured metabolites that were shown to be affected by the meal consumed included the essential amino acids L.Histidine, L.Isoleucine, L.Leucine, L.Lysine, L.Methionine, L.Phenylalanine, and L.Tryptophan. 

The first principal component is a good summary of the change in plasma
levels of ...

-   L.Lysine as it accounts for 0.96 of its variability

-   L.Tryptophan as it accounts for 0.91 of its variability

-   L.Methionine as it accounts for 0.86 of its variability

The first principal component is not the best summary of a change in
plasma levels of ...

-   L.Isoleucine as it accounts for 0.76 of its variability

-   L-Histidine as it accounts for 0.73 of its variability

The first principal component is not a very good summary of a change in
plasma levels of ...

-   L.Phenylalanine as it accounts for 0.41 of its variability

-   L.Leucine as it accounts for 0.53 of its variability

Principal components 2 and 3 may want to be investigated to explore the variability of the change in plasma levels for these metabolites.
    
> Make a plot of the first principal component score vs time, using a separate panel for each meal. 
> Summarize the difference in patterns between the meals.

```{r}
library(ggplot2)

data1<-read.csv("metab.selected.csv")
PCA$x[,1]->PCA1

# Create a separate plot for each meal using facet_wrap
ggplot(data1, aes(x=Time, y=PCA1, color=Meal)) + 
  geom_point() +
  facet_wrap(vars(Meal)) +
  xlab("Time") +
  ylab("PC1 Score") +
  ggtitle("PC1 Score vs Time by Meal")
```

Meal A has the lowest range of variability from about -2 to 5, whereas
Meal C has the largest range of variability from about -15 to 5. There
is one outliier within Meal A which could be investigated further. Meal
A PC1 scores has a relatively linear trend throughout the whole
250minutes. In contrast, Meal B PC1 scores have an initial decreasing
trend until around 75 minutes, where they begin to increase again until
250 minutes. Meal C has the largest range of where the scores decreasing
trend until about 125 minutes, where the scores begin to increase again
until 250 minutes.

> Q5. There is one outlier at time 0, meal A (the control meal). What participant does this measurement corresponds to?

```{r}
# Subset data to only include rows where Time is 0
library(tibble)
library(dplyr)

d1<-data1 %>% 
  mutate(PC1 = PCA1) %>%
  filter(Time == 0, Meal == "A")

which.min(d1$PC1)
d1$Participant[24]
```

The outlier at time zero, Meal A (the control meal) corresponds LD4227.

> 6. Make a plot of the time course of score 1 for each meal, faceted by individual.
> Based on this plot, comment on whether the outlier from (5) represents a) a weird observation; b) a weird day (the whole time course for that meal is unusual); or c) a weird participant. 

This outlier from (5) could represent a weird participant. Those in Meal
A were given a control (water), so the results from that experiment show
participant's baseline blood metabolite levels (assuming they haven't
eaten anything else).

The outlier at LD4227 consistently displays an unusual deviation from
the rest of the participants blood plasma levels across other Meal
types. It is plausible that the outlier in (5) does not consume enough
protein in their general diet, causing their blood metabolite levels to
be extremely low.

```{r}

ggplot(data1, aes(x = Time, y = PCA1, color = Meal)) +
  geom_line() +
  facet_wrap(vars(Participant)) +
  labs(x = "Time", y = "Score 1")
```
