---
title: "Multivariate Analysis on Metabolites"
author: "Brittany Alexandra"
date: "Wednesday 15 March 2023`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggfortify)
library(readr)
library(FactoMineR)
library(factoextra)
library(gridExtra)
library(tibble)
library(knitr)
```

# Introduction

This is an exploratory analysis of metabolites affected by whey
consumption utilizing a relatively clean dataset. As someone with a love
for biology and the gym, this was a fascinating dataset to analyse, and
a great test on my understanding of multivariate analysis techniques.

# Findings

Certain metabolites contributed significantly to the variance in this
analysis, indicating their significant reactions to whey consumption.
These included L.Histidine, L.Isoleucine, L.Leucine, L.Lysine,
L.Methionine, L.Phenylalanine, and L.Tryptophan.

Each of these Amino Acids participate in metabolic pathways. Therefore,
understanding how they are influenced provides a foundation for further
investigation in how to enhance, or reduce these Amino Acids and their
consequent biological consequences.

For example;

-   **L.Leucine** is renowned for its ability to activate mTOR
    signaling, and is a central regulator of muscle protein synthesis.
    It's involvement in this critical pathway underscores its
    significance in the context of muscle growth and repair.

-   **L.Tryptophan** serves as a precursor to serotonin, which is a
    neurotransmitter crucial for mood regulation. L.Tryptophan wields
    influence over our emotional well-being, and understanding how its
    levels respond to whey protein consumption opens a window into
    potential therapeutic avenues aimed at elevating serotonin levels
    for mood enhancement.

-   **L.Lysine** is involved in collagen formation, calcium absorption,
    and the synthesis of carnitine, a molecule crucial for energy
    production.Therefore, monitoring L.Lysine levels in response to whey
    protein consumption can help understand its impact on skin health,
    bone health, and energy metabolism.

# Concerns 

1.  This is a snapshot of data and therefore does not capture the
    inherent individual variability
2.  This snapshot of data was done in an isolated environment, and would
    therefore be inaccurate to generalize to a wider population. Further
    tests with a more diverse sample size representative of the
    population would be required to generalise these results.
3.  Limited time-frame thereby not accounting for variability into
    metabolism and digestion over longer periods of time i.e. \>24
    hours.

# Background

The "metab-selected.csv" file presents an intriguing snapshot from a
study on whey protein beverages and their impact on blood metabolites
(as detailed in Joblin-Mills et al., Frontiers in Clinical Diabetes).

It captures the dynamic response of 52 significant metabolites to three
beverage variants; a water-based control, a standard whey protein drink,
and a double-concentration whey variant.

These were tracked over key time intervals post-consumption. This
dataset, with its log-transformed values, offers a window into
understanding how dietary changes can distinctly alter metabolic
profiles.

# Statistical Analysis

```{r}
# Lets take a look at the data to get an understanding of it
data <- read_csv("~/Desktop/metab-selected.csv")
head(data)
```

## 1. Lets reduce the dimensional of our data to understand the structure and covariance of metabolites

```{r}

#Clean data
data<-data[,-c(1:3)]
data<-log(data)
missing<-apply(is.na(data),2 , sum)
data<-data[,missing==0]
sd<-apply(data, 2, sd)
data<-data[,sd>0.001]


#Data matrix for simple, effective, but ugly plot 
PCA<-prcomp(data, center=T, scale = T)
# plot(PCA)

# Limit the data to the first 8 principal components to not overwhelm 8 axis 
variance_explained <- PCA$sdev^2 / sum(PCA$sdev^2)  # Proportion of variance explained by each PC

# Create a dataframe with Principal Component and Variance Explained
variance_df <- data.frame(
  PrincipalComponent = 1:length(variance_explained),
  VarianceExplained = variance_explained
)

variance_df <- variance_df[1:8,]

# Create a bar chart for the first 6 principal components
gg_var_plot <- ggplot(variance_df, aes(x = PrincipalComponent, y = VarianceExplained)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Variance Explained by the First 6 Principal Components",
       subtitle = "Plot 1",
       x = "Principal Component",
       y = "Proportion of Variance Explained") +
  geom_text(aes(label = scales::percent(VarianceExplained, accuracy = 1)), vjust = -0.5)

# Print the bar chart
print(gg_var_plot)

# Assuming PCA is your PCA result object
loadings <- as.data.frame(PCA$rotation[, 1])  # Extract loadings for PC1
names(loadings) <- "Contribution"
loadings$Variable <- rownames(loadings)

# Create the bar plot
ggplot(loadings, aes(x = reorder(Variable, Contribution), y = Contribution)) +
  geom_bar(stat = "identity", col = "gray", position = position_dodge(width = 0.7), width = 0.7) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Contribution of Variables to PC1",
       subtitle = "Plot 2",
       x = "Variable",
       y = "Contribution") +
  geom_hline(yintercept = c(-0.2, 0.2), linetype = "dashed", color = c("red", "red"), size = 1)
  #annotate("text", x = 1.5, y = 0.25, label = "Metabolites beyond -0.2 and 0.2 contribute significantly", color = "black", size = 4)


```

Overall, we can understand a large amount of variation from the first
Principal Component (Plot 1). (PC summarizes the core differences or
similarities within the data). This gives us a benchmark for
investigating dominant underlying factors that are influencing the
observed measurements.

Metabolites beyond -0.2 and 0.2 contribute significantly to the variance
(Plot 2).

For further studies, isolating the top 10 metabolites and see how they
react to whey consumption and the effects on the human body would be
interesting.

For instance, we can see Lysine contributes the most with \~0.25 of the
variance coming from this metabolite. However, it is not surprising that
lysine would contribute significantly to variability in a study
involving metabolites. Lysine, being an essential Amino Aic, can only be
obtained through dietary sources, and its availability in the body is
directly influenced by dietary intake.

**Notes**

28.6% of the variability in blood metabolites following Whey Protein
consumption is accounted for by the first principal component, and 9.5%
for the second PC.

What does this mean? Well, in PCA, each principal component represents a
linear combination of the original variables. The first PC captures the
most significant patterns of variation in the data, while the subsequent
PCs capture the remaining variability in decreasing order of importance.

As the first principal component explains 28.6% of the total
variability, we can say it is responsible for a substantial portion of
the patterns observed in the blood metabolites following Whey Protein
consumption. This component likely represents the primary factor
influencing the changes in metabolite levels.

Similarly, the second principal component accounts for 9.5% of the
variability. While it explains less variability compared to the first
PC, it still captures a notable proportion of the patterns in the data.
The second component likely represents additional factors that
contribute to the variability in blood metabolites, possibly
representing specific metabolic responses or interactions.

## What variables were signifcantly affected by the meals post whey consumption?

By understanding the variables that have a strong correlation with the
first PC score, we can identify the metabolites most significant
imapcted by whey consumption are:

L.Histidine, L.Isoleucine, L.Leucine, L.Lysine, L.Methionine,
L.Phenylalanine, and L.Tryptophan.

```{r}
# Coefficients and their strength with PC1
coef1 <- round(PCA$rotation[, 1] * PCA$sdev[1], 2)
variable_names <- names(coef1)

coef_table <- data.frame(Variable = variable_names) # Create a data frame with a single "Variable" column
coef_table$Coefficient <- coef1 # Create a new column for the coefficients
sorted_coef_table <- coef_table[order(coef1, decreasing = TRUE), ] # Sort the table based on the coefficients

# Remove the row numbers (index)
rownames(sorted_coef_table) <- NULL

# Create a nice table
kable(sorted_coef_table, 
      caption = "Coefficients and Their Strength with PC1",
      align = "c",  
      col.names = c("Variable", "Coefficient"),
      format.args = list(big.mark = ",", scientific = FALSE),  # Formatting options
      booktabs = TRUE)

```

**Notes**

How was this identified?

Using the rotation matrix and the standard deviation associated with
principal component 1, I computed the correlation of the first principal
component score with the original variables.

The plot helps you see which original variables have the most impact on
PC1 (those far to the right) and how their impact relates to how much
they vary (higher or lower on the plot). It helps identify which
variables are driving the patterns captured by PC1.

Positive correlation: A positive correlation indicates that as the PC1
score increases, the corresponding metabolite tends to increase as well.
These metabolites are positively associated with PC1 and may contribute
to its variation.

Negative correlation: A negative correlation suggests that as the PC1
score increases, the corresponding metabolite tends to decrease. These
metabolites are negatively associated with PC1 and may have an inverse
relationship with its variation.

Therefore :

The first principal component is a good summary of the change in plasma
levels of ...

-   L.Lysine as it accounts for 0.96 of its variability

-   L.Tryptophan as it accounts for 0.91 of its variability

-   L.Methionine as it accounts for 0.86 of its variability

The first principal component is not the best summary of a change in
plasma levels of ...

-   L.Isoleucine as it accounts for 0.76 of its variability

-   L-Histidine as it accounts for 0.73 of its variability

The first principal component is not a very good summary of a change in
plasma levels of ...

-   L.Phenylalanine as it accounts for 0.41 of its variability

-   L.Leucine as it accounts for 0.53 of its variability

Principal components 2 and 3 may want to be investigated to explore the
variability of the change in plasma levels for these metabolites.

## What are the key difference in patterns between meals?

```{r}
# Make a plot of the first principal component score vs time, using a separate panel for each meal.
data1 <- read_csv("~/Desktop/metab-selected.csv")
PCA$x[,1]->PCA1

# Create a separate plot for each meal using facet_wrap
ggplot(data1, aes(x=Time, y=PCA1, color=Meal)) + 
  geom_point() +
  facet_wrap(vars(Meal)) +
  xlab("Time") +
  ylab("PC1 Score") +
  ggtitle("PC1 Score vs Time by Meal")
```

-   Meal A has the lowest range of variability from about -2 to 5,
    whereas Meal C has the largest range of variability from about -15
    to 5.

-   There is one outliier within Meal A which could be investigated
    further. Meal A PC1 scores has a relatively linear trend throughout
    the whole 250minutes.

-   In contrast, Meal B PC1 scores have an initial decreasing trend
    until around 75 minutes, where they begin to increase again until
    250 minutes.

-   Meal C has the largest range of where the scores decreasing trend
    until about 125 minutes, where the scores begin to increase again
    until 250 minutes.

## Lets investigate the outlier at time 0, meal A (the control meal).

```{r}
# Subset data to only include rows where Time is 0
d1<-data1 %>% 
  mutate(PC1 = PCA1) %>%
  filter(Time == 0, Meal == "A")

which.min(d1$PC1)
d1$Participant[24]
```

The outlier at time zero, Meal A (the control meal) corresponds LD4227.

```{r}
ggplot(data1, aes(x = Time, y = PCA1, color = Meal)) +
  geom_line() +
  facet_wrap(vars(Participant)) +
  labs(x = "Time", y = "Score 1")
```

It is likely they are either...\
a) a weird observation; b) a weird day (the whole time course for that
meal is unusual); or c) a weird participant.

Its likely the outlier from (5) could represent a weird participant.
Those in Meal A were given a control (water), so the results from that
experiment show participant's baseline blood metabolite levels (assuming
they haven't eaten anything else).

The outlier at LD4227 consistently displays an unusual deviation from
the rest of the participants blood plasma levels across other Meal
types. It is plausible that the outlier in (5) does not consume enough
protein in their general diet, causing their blood metabolite levels to
be extremely low.
